A More Complete Algebra for File Synchronization

### 2015-11-08 2015/paper.tex
Separated subgroups of axioms

\newcommand{\eqext}{\sqsubseteq}
\newcommand{\eqnrw}{\sqsupseteq}

@@

\title{A Complete Algebra for File Synchronization}

@@

% diagram
% /> D ---> F ---> BOT <\
% \--/               \--/
\input{type-transition}

@@

We also know that after $create$ or $edit$, $\FS(p)\neq\empt$, whereas after $remove$,
$\FS(p)$ will be $\empt$. However, from \cite{NREC:alg} we know that a useful set of axioms
will need to distinguish between edits that result in directories ($edit(p,D)$) and
ones that result in files ($edit(p,F)$). This creates a seemingly arbitrary asymmetry where
there is more information encoded into commands about their results than about the
original state of the filesystem.

@@

We aim to use our algebra to define and implement algorithms for conflict detection
and ordering updates (commands) before we can apply them to all replicas.
Conflicts and ordering all happen between commands, therefore in our algebra
we will reason about sequences of commands, 
and do so based on a set of axioms about pairs of neighbouring commands.
We are interested to see what pairs of commands can be simplified or reversed
in order to see under what circumstances commands conflict or can be reordered.

Pairs of commands in general will have the form
\[ [\cxy(p,V_Y); \czw(q,V_W)] \]
where $X,Y,Z,W\in\{\fsregcommandchar{D},\fsregcommandchar{F},\empt\}$, $p,q\in\setp$, and values are of the appropriate type: 
$V_Y\in\setvx{Y}$ and $V_W\in\setvx{W}$
(although commands where $Y=\empt$ or $W=\empt$ do not require a value).

Similarly to the case of commands, in \cite{NREC:alg} we found that in order to build
a set of axioms in a sound and complete algebra we at times needed to take into account
the relationship between the paths of the commands in the pair, $p$ and $q$.
As the relationship between the paths determine how the commands affect the filesystem
and its tree property, 
and again in line with the aim of providing as much information as possible for the algebra,
we encode this relationship in the pair of commands. The possible relationships we
consider are as follows.

\begin{itemize}
\item $p=q$. If the two commands affect the same path, the pair of commands
will clearly behave differently than if the paths are different.
%
\item Separate paths, that is, $\parent(p)\neq q$ and $\parent(q)\neq p$. If the paths
are not directly related, then modifying the filesystem at one path is not going
to affect the outcome of the command on the other path.
The two commands will therefore commute and the pair will be reversible.
We will denote a path separate from $p$ with $\np$.
%
\item Directly related paths, that is, $\parent(p)=q$ or $\parent(q)=p$.
For simplicity, we will write $\pp$ for $\parent(p)$.
Some of these command pairs will not be reversible.
For example, $[\cbd(\pp,D);\cbf(p,F)]$ will work on some filesystems,
while $[\cbf(p,F);\cbd(\pp,D)]$ will break all filesystems.
\end{itemize}

@@

In summary, we have pairs of commands
of the following forms:
\begin{itemize}
\item[i.] \( [\cxy(p,V_Y); \czw(\np,V_W)] \)
\item[ii.] \( [\cxy(p,V_Y); \czw(p,V_W)] \)
\item[iii.] \( [\cxy(\pp,V_Y); \czw(p,V_W)] \)
\item[iv.] \( [\cxy(p,V_Y); \czw(\pp,V_W)] \)
\end{itemize}

@@

\begin{itemize}
\item[i.] If the two paths in the pair are separate (not directly related), then
the commands commute with an equivalence:
\[ [\cxy(p,V_Y); \czw(\np,V_W)] \equiv [\czw(\np,V_W); \cxy(p,V_Y)] \]
%
\item[ii.] If the two paths are the same ($[\cxy(p,V_Y); \czw(p,V_W)]$), then
the pair will break every filesystem if $Z\ne Y$, that is, if the second
command expects a type of value different from the one left by the first command.
Otherwise, the pair will simplify to a single or no commands
if both $X$ and $W$ are $\fsregcommandchar{\empt}$ or $\fsregcommandchar{D}$:
\begin{gather*}
% bf=fb Sp
           [\cbf(p, F); \cfb(p)] \eqext \emptyseq \\
% bd=db Sp
           [\cbd(p, D); \cdb(p)] \eqext \emptyseq \\
% db=bd Sp
           [\cdb(p); \cbd(p, D)] \eqext [\cdd(p, D)] \\
% df=fd Sp
           [\cdf(p, F); \cfd(p, D)] \eqext [\cdd(p, D)]
\end{gather*}
In all other cases the commands are equivalent to a single command
replacing the input of the first command with the output of the second:
\[ [\cxy(p, V_Y); \cyz(p, V_Z)] \equiv [\cxz(p, V_Z)] \]
%
\item[iii.]
Of the pairs in which the first command is applied to the parent of the path
of the second command there are pairs to which no law applies.
We call these ``construction'' pairs:
\begin{gather*}
% bd>bf
           [\cbd(\pp, D); \cbf(p, F)] \\
% bd>bd
           [\cbd(\pp, D_1); \cbd(p, D_2)] \\
% fd>bf
           [\cfd(\pp, D); \cbf(p, F)] \\
% fd>bd
           [\cfd(\pp, D_1); \cbd(p, D_2)]
\end{gather*}
Otherwise commands in these pairs commute iff the first command is $\cdd$;
all other pairs break every filesystem. For example:
\begin{gather*}
[\cdd(\pp, D); \cbf(p, F)] \equiv [\cbf(p, F); \cdd(\pp, D)] \\
[\cdf(\pp, F); \cbd(p, D)] \equiv \cbrk
\end{gather*}
%
\item[iv.]
The final group of pairs are the ones in which the path in the second
command is the parent of the one in the first.
These behave similarly to the previous group. There are
so-called ``destruction'' pairs to which no law applies:
\begin{gather*}
% fb<db ??
           [\cfb(p); \cdb(\pp)] \\
% fb<df ??
           [\cfb(p); \cdf(\pp, F)] \\
% db<db ??
           [\cdb(p); \cdb(\pp)] \\
% db<df ??
           [\cdb(p); \cdf(\pp, F)]
\end{gather*}
Otherwise the commands commute iff the \emph{second} command is $\cdd$.
All other pairs break every filesystem.
\end{itemize}

@@

\begin{itemize}
\item[i.] Pairs with separate paths are still reversible,
but \emph{these are the only pairs that are so}.
%
\item[ii.] If the two paths are the same ($[\cxy(p,V_Y); \czw(p,V_W)]$), then, as above,
the pair will break every filesystem if $Z\ne Y$.
Otherwise, the pair is extended by an empty sequence
if both $X$ and $W$ are $\fsregcommandchar{\empt}$ or $\fsregcommandchar{D}$:
\begin{gather*}
% bf=fb Sp
           [\cbf(p, F); \cfb(p)] \eqext \emptyseq \\
% bd=db Sp
           [\cbd(p, D); \cdb(p)] \eqext \emptyseq \\
% db=bd Sp
           [\cdb(p); \cbd(p, D)] \eqext \emptyseq \\
% df=fd Sp
           [\cdf(p, F); \cfd(p, D)] \eqext \emptyseq
\end{gather*}
And in all other cases, as above, the commands are equivalent to a single command
replacing the input of the first command with the output of the second:
\[ [\cxy(p, V_Y); \cyz(p, V_Z)] \equiv [\cxz(p, V_Z)] \]
%
\item[iii.]
Of the pairs in which the first command is applied to the parent path,
the same ``construction'' pairs still have no laws, but all other pairs
break all filesystems.
%
\item[iv.]
And we still have the ``destruction'' pairs among the pairs where the second command
is applied to the parent path, and all other pairs break all filesystems.
\end{itemize}

@@

\bibitem{BZ} Bill Zissimopoulos, ...

>>>>

\newcommand{\eqext}{\sqsubseteq}
\newcommand{\eqnrw}{\sqsupseteq}
\newcommand{\nequiv}{\not\equiv}

@@

\title{A More Complete Algebra for File Synchronization}

@@

\input{type-transition}

@@

We also know that after $create$ or $edit$, $\FS(p)\neq\empt$, whereas after $remove$,
$\FS(p)$ will be $\empt$. However, from \cite{NREC:alg} we know that a useful set of axioms
will in some cases need to distinguish between edits that result in directories ($edit(p,D)$) and
ones that result in files ($edit(p,F)$);
and Bill Zissimopoulos suggested in \cite{BZ} that extending this distinction simplifies
definitions based on our algebra, as it is then able to predict the behaviour of commands
more precisely. 
In other words, encoding this information in the commands is definitely useful,
but it creates a seemingly arbitrary asymmetry where
there is more information encoded into commands about their results than about the
original state of the filesystem.

@@

We aim to use our algebra to define and implement algorithms for conflict detection
and ordering updates (commands) before we can apply them to all replicas.
Conflicts and ordering all happen between commands, therefore in our algebra
we will reason about sequences of commands, 
and do so based on a set of axioms about pairs of neighbouring commands.
We are interested to see which pairs of commands cause errors all the time,
and which can be simplified or reversed
in order to see under what circumstances commands conflict or can be reordered.

Pairs of commands in general will have the form
\[ [\cxy(p,V_Y); \czw(q,V_W)] \]
where $X,Y,Z,W\in\{\fsregcommandchar{D},\fsregcommandchar{F},\empt\}$, $p,q\in\setp$, and values are of the appropriate type: 
$V_Y\in\setvx{Y}$ and $V_W\in\setvx{W}$.
(Strictly speaking, commands where $Y=\empt$ or $W=\empt$ do not require a value,
but they can also be written in a $\fscommand{X}{\empt}(p, \empt)$ form.)

Similarly to the case of commands, in \cite{NREC:alg} we found that in order to build
a set of axioms in a sound and complete algebra and use the algebra effectively we at times needed to take into account
the relationship between the paths of the commands in the pair, $p$ and $q$.
As the relationship between the paths determine how the commands affect the filesystem
and its tree property, 
and again in line with the aim of providing as much information as possible for the algebra,
we encode this relationship in the pair of commands. The possible relationships we
consider are as follows.

\begin{description}
\item[Same.] Paths that are the same ($p=q$). If the two commands affect the same path, the pair of commands
will clearly behave differently than if the paths are different.
%
\item[Unrelated.] Paths that are not \emph{directly} related,
that is, where $\parent(p)\neq q$ and $\parent(q)\neq p$. 
In this case modifying the filesystem at one path is not going
to affect the outcome of the command on the other path.
The two commands will therefore commute and the pair will be reversible.
We will denote a path separate from $p$ with $\np$.
%
\item[Directly related.] Directly related paths, that is, $\parent(p)=q$ or $\parent(q)=p$.
For simplicity, we will write $\pp$ for $\parent(p)$.
Some of these command pairs will not be reversible.
For example, $[\cbd(\pp,D);\cbf(p,F)]$ will work on some filesystems,
while $[\cbf(p,F);\cbd(\pp,D)]$ will break all filesystems.
\end{description}

@@

% TODO Would it be worth encoding non-direct parent/child relationships?

In summary, we have pairs of commands
of the following forms:
\begin{description}
\item[Unrelated.] \( [\cxy(p,V_Y); \czw(\np,V_W)] \)
\item[Same.] \( [\cxy(p,V_Y); \czw(p,V_W)] \)
\item[Down.] \( [\cxy(\pp,V_Y); \czw(p,V_W)] \)
\item[Up.] \( [\cxy(p,V_Y); \czw(\pp,V_W)] \)
\end{description}

@@

\begin{description}
\item[Unrelated.] If the two paths in the pair are not directly related, then
the commands commute with an equivalence and they will not break every filesystem:
\begin{align*}
[\cxy(p,V_Y); \czw(\np,V_W)] &\equiv [\czw(\np,V_W); \cxy(p,V_Y)] \\
&\nequiv \cbrk
\end{align*}
%
\item[Same.] If the two paths are the same ($[\cxy(p,V_Y); \czw(p,V_W)]$), then:
\begin{description}
\item[Break.]
The pair will break every filesystem if $Z\ne Y$, that is, if the second
command expects a type of value different from the one left by the first command.
\item[Simplified by extension.]
Otherwise, the pair will simplify to a single or no commands
if both $X$ and $W$ are $\fsregcommandchar{\empt}$ or $\fsregcommandchar{D}$:
\begin{gather*}
% bf=fb Sp
           [\cbf(p, F); \cfb(p)] \eqext \emptyseq \\
% bd=db Sp
           [\cbd(p, D); \cdb(p)] \eqext \emptyseq \\
% db=bd Sp
           [\cdb(p); \cbd(p, D)] \eqext [\cdd(p, D)] \\
% df=fd Sp
           [\cdf(p, F); \cfd(p, D)] \eqext [\cdd(p, D)]
\end{gather*}
\item[Simplified by equivalence.]
In all other cases, the commands are equivalent to a single command
replacing the input of the first command with the output of the second:
\[ [\cxy(p, V_Y); \cyz(p, V_Z)] \equiv [\cxz(p, V_Z)] \]
\end{description}
%
\item[Down.]
These are the pairs in which the first command is applied to the parent of the path
of the second command.
\begin{description}
\item[Construction.] There are pairs in this group to which no law applies.
We call these ``construction'' pairs:
\begin{gather*}
% bd>bf
           [\cbd(\pp, D); \cbf(p, F)] \\
% bd>bd
           [\cbd(\pp, D_1); \cbd(p, D_2)] \\
% fd>bf
           [\cfd(\pp, D); \cbf(p, F)] \\
% fd>bd
           [\cfd(\pp, D_1); \cbd(p, D_2)]
\end{gather*}
\item[Commute.] Otherwise, the commands in these pairs commute iff the first command is $\cdd$. For example:
\begin{gather*}
[\cdd(\pp, D); \cbf(p, F)] \equiv [\cbf(p, F); \cdd(\pp, D)]
\end{gather*}
\item[Break.] All other pairs break every filesystem, e.g.:
\begin{gather*}
[\cdf(\pp, F); \cbd(p, D)] \equiv \cbrk
\end{gather*}
\end{description}
%
\item[Up.]
The final group of pairs are the ones in which the path in the second
command is the parent of the one in the first.
These behave similarly to the previous group. 
\begin{description}
\item[Desctruction.] There are
so-called ``destruction'' pairs to which no law applies:
\begin{gather*}
% fb<db ??
           [\cfb(p); \cdb(\pp)] \\
% fb<df ??
           [\cfb(p); \cdf(\pp, F)] \\
% db<db ??
           [\cdb(p); \cdb(\pp)] \\
% db<df ??
           [\cdb(p); \cdf(\pp, F)]
\end{gather*}
\item[Commute.] Otherwise the commands commute iff the \emph{second} command is $\cdd$.
\item[Break.] All other pairs break every filesystem.
\end{description}
\end{description}

@@

\begin{description}
\item[Unrelated.] Pairs with unrelated paths are still reversible and not equivalent to
$\cbrk$, but these are the only pairs that are reversible.
%
\item[Same.] If the two paths are the same ($[\cxy(p,V_Y); \czw(p,V_W)]$), then:
\begin{description}
\item[Break.] As above, the pair will break every filesystem if $Z\ne Y$.
\item[Simplified by extension.] Otherwise, the pair is extended by an empty sequence
if both $X$ and $W$ are $\fsregcommandchar{\empt}$ or $\fsregcommandchar{D}$:
\begin{gather*}
% bf=fb Sp
           [\cbf(p, F); \cfb(p)] \eqext \emptyseq \\
% bd=db Sp
           [\cbd(p, D); \cdb(p)] \eqext \emptyseq \\
% db=bd Sp
           [\cdb(p); \cbd(p, D)] \eqext \emptyseq \\
% df=fd Sp
           [\cdf(p, F); \cfd(p, D)] \eqext \emptyseq
\end{gather*}
\item[Simplified by equivalence.]
And in all other cases, as above, the commands are equivalent to a single command
replacing the input of the first command with the output of the second:
\[ [\cxy(p, V_Y); \cyz(p, V_Z)] \equiv [\cxz(p, V_Z)] \]
\end{description}
%
\item[Down.]
Of the pairs in which the first command is applied to the parent path:
\begin{description}
\item[Construction.] the same ``construction'' pairs still have no laws, 
\item[Break.] but all other pairs break all filesystems.
\end{description}
%
\item[iv.]
And of the pairs where the second command
is applied to the parent path,
\begin{description}
\item[Destruction.] we still have the ``destruction'' pairs,
\item[Break.] and all other pairs break all filesystems.
\end{description}
\end{description}

@@

\bibitem{BZ} Bill Zissimopoulos, personal communication, ...
